### 2015-03-17 doc/thesis/typedlua.tex
Minor improvements to the last part of Chapter 3

\begin{figure}[!ht]
\textbf{Classes}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:classes}
\end{figure}

@@

Lua has a mechanism for self-like (or JavaScript-like) delegation of
missing fields in a table.
After the call \texttt{setmetatable(t1, \{ \string_\string_index = t2 \})},
Lua looks up in \texttt{t2} for any missing fields of \texttt{t1}.
As we mentioned in Section \ref{sec:statistics},
Lua programmers often use this mechanism to simulate classes.

Typed Lua combines this mechanism along with type aliases and
interfaces to allow programmers to define classes in a similar
fashion to the definition of modules.
We use the following example to discuss how Typed Lua type checks
object-oriented code:
\begin{verbatim}
    interface Shape
      x, y:number
      const new:(number, number) => (self)
      const move:(number, number) => ()
    end
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:new (x:number, y:number)
      local s = setmetatable({}, { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move (dx:number, dy:number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    return Shape
\end{verbatim}

In this example, we use an interface to define the type of the class \texttt{Shape}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types work.
We use a double arrow instead of a single arrow because it is
syntactic sugar for defining the type of methods,
as it includes an implicit first type \texttt{self}.
The double arrow can appear only inside the declaration of
interfaces and userdata, which we will introduce in the next section.
This interface definition exports the following object type to the class \texttt{Shape}:
\begin{verbatim}
    { "x":number, "y":number,
      const "new":(self, number, number) -> (self),
      const "move":(self, number, number) -> () }
\end{verbatim}

In Typed Lua, \texttt{setmetatable} is a strict primitive.
Typed Lua only allows calling \texttt{setmetatable} with
two operands.
The first operand can be any table \texttt{t1}, but
the second operand must be a table of the form
\texttt{\{ \string_\string_index = t2 \}},
where \texttt{t2} can be any table.
The reason for being so strict with \texttt{setmetatable} is because
it is the only mechanism that Typed Lua has to create classes
as prototype objects.
In a \texttt{setmetatable} expression, if the type of \texttt{t2}
is a subtype of the type of \texttt{t1}, it changes \texttt{t1} to
have the same type of \texttt{t2}.

In the previous example, we use \texttt{setmetatable} to define a constructor
\texttt{new} to the class \texttt{Shape}.
Note that we use \texttt{setmetatable} inside \texttt{new} to declare
the local \texttt{s} having the same type of \texttt{self}.

Still in the previous example, we are using the refinement of table
types to build the type of the variable \texttt{Shape},
as it should work as our object prototype, matching the type
that the interface declaration exports.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
Finally, the local \texttt{Shape} has the same type of the
interface declaration in the last \texttt{return} statement.

In Typed Lua, the style of classes definition from the previous example
allow us to use \texttt{require} for creating prototype objects that
work as classes, along with an alias to the table type of the class.
This allows us to use the exported alias in type annotations,
as we show in the following example:
\begin{verbatim}
    local Shape = require "shape"
    local shape1 = Shape:new(0, 5)
    local shape2:Shape = Shape:new(10, 10)
\end{verbatim}

Typed Lua also uses \texttt{setmetatable} to simulate single inheritance.
We use the following example to discuss single inheritance in Typed Lua:
\begin{verbatim}
    local Shape = require "shape"
    interface Circle
      x, y, radius:number
      const new:(number, number, value) => (self)
      const move:(number, number) => ()
      const area:() => (number)
    end
    local Circle = setmetatable({}, { __index = Shape })
    Circle.radius = 0.0
    const function Circle:new (x:number, y:number, radius:value)
      local c = setmetatable(Shape:new(x, y), { __index = self })
      c.radius = tonumber(radius) or 0.0
      return c
    end
    const function Circle:area ()
      return math.pi * self.radius * self.radius
    end
    return Circle
\end{verbatim}

In this example, we use an interface to define the type of the class \texttt{Circle},
but with a syntactical overhead to manually include new and
overridden methods.
This interface definition exports the following object type to the class \texttt{Circle}:
\begin{verbatim}
    { "x":number, "y":number, "radius":number,
      const "new":(self, number, number, value) -> (self),
      const "move":(self, number, number) -> (),
      const "radius":() -> (number) }
\end{verbatim}

In the previous example, we use \texttt{setmetatable} in two different ways.
In the first case, we use \texttt{setmetatable} to trigger the refinement
of table types, as we need it to add new methods and override existing ones
in \texttt{Circle}.
This is possible because a call to \texttt{setmetatable},
where \texttt{t1} is a table constructor assigned to a fresh local variable,
and \texttt{t2} is not \texttt{self} gives this variable an open type.
In the second case, we use \texttt{setmetatable} to redefine \texttt{new}.
Note how we can call the constructor of \texttt{Shape} inside
the overridden constructor.
A limitation of this class system is that the overridden constructor
must be a subtype of the original constructor,
so the type of \texttt{radius} has to be quite permissive.

@@

In all examples, if we erase all type and \texttt{const} annotations,
and type declarations, they become valid Lua code, which run with the
same semantics as Typed Lua code.

>>>>

@@

Lua has a mechanism for self-like (or JavaScript-like) delegation of
missing fields in a table.
After the call \texttt{setmetatable(t1, \{ \string_\string_index = t2 \})},
Lua looks up in \texttt{t2} for any missing fields of \texttt{t1}.
As we mentioned in Section \ref{sec:statistics},
Lua programmers often use this mechanism to simulate classes.
We will use the following example to discuss this feature:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:new (x:number, y:number):self
      local s:self = setmetatable({}, { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move (dx:number, dy:number):()
      self.x = self.x + dx
      self.y = self.y + dy
    end
    return Shape
\end{verbatim}

In Typed Lua, \texttt{setmetatable} is a strict primitive
that obeys three typing rules.
The reason for being so strict with \texttt{setmetatable} is because
it is the only mechanism that Typed Lua has to create classes
as prototype objects.

The first \texttt{setmetatable} rule appears in our previous example.
In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has type \texttt{self} then the expression also also has type \texttt{self}.
We will explain the other two rules while we explain how
Typed Lua type checks single inheritance.

In our example, we are using the refinement of table types to
build the type of the variable \texttt{Shape},
as it should work as our class.
This means that the local \texttt{Shape} has type
\texttt{\{"x":number, "y":number\}} inside the definition of \texttt{new}.
This also makes Typed Lua bind the type of the local \texttt{Shape}
to the type \texttt{self} inside the definition of \texttt{new},
allowing us to access fields \texttt{x} and \texttt{y}.
After the definition of \texttt{new}, the local \texttt{Shape}
has type
\texttt{\{"x":number, "y":number, const "new":(self, number, number) -> (self)\}},
which is the type that Typed Lua binds to the type \texttt{self}
inside the definition of \texttt{move}.
We use the \texttt{const} annotation in the type of the methods
because it is necessary for covariance among object types work.
Finally, the last \texttt{return} statement fixes the type of
\texttt{Shape} because \emph{fixed} table types do not allow width
subtyping, and this behavior allows Typed Lua to use the refinment
of table types to type check single inheritance, as we will see in
this section.

A return statement that appears in the top-level also exports
a type alias to be used in type annotations.
For instance, our example exports the following type alias:
\begin{verbatim}
    typealias Shape =
      <{ "x":number, "y":number,
         const "new":(self, number, number) -> (self),
         const "move":(self, number, number) -> () }>
\end{verbatim}

The style of classes definition from the previous example
allow us to use \texttt{require} for creating prototype objects that
work as classes, along with an alias to the table type of the class.
This allows us to use the exported alias in type annotations,
as we show in the following example:
\begin{verbatim}
    local Shape = require "shape"
    local shape1 = Shape:new(0, 5)
    local shape2:Shape = Shape:new(10, 10)
\end{verbatim}

Typed Lua also includes a \texttt{class} construct that programmers
can use to document the type of the classes they export..
Figure \ref{fig:classes} shows the concrete syntax of the
\texttt{class} construct.

\begin{figure}[!ht]
\textbf{Classes}\\
\dstart
\begin{align*}
\textit{class} ::= & \;\; [\textbf{local}] \; \textbf{class} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; \{\textit{decitem}\} \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}\\
\textit{methodtype} ::= & \;\; \textit{tupletype} \; \texttt{`=>'} \; \textit{rettype}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua classes}
\label{fig:classes}
\end{figure}

As an example, we can use the following \texttt{class} declaration to
document the type \texttt{Shape} from the previous example:
\begin{verbatim}
    class Shape
      x, y:number
      const new:(number, number) => (self)
      const move:(number, number) => ()
    end
\end{verbatim}

We use a double arrow instead of a single arrow because it is
syntactic sugar for defining the type of methods,
as it includes an implicit first type \texttt{self}.
The double arrow can appear only inside the declaration of
classes and userdata, which we will introduce in the next section.

Typed Lua also uses \texttt{setmetatable} to simulate single inheritance.
We use the following example to discuss single inheritance in Typed Lua:
\begin{verbatim}
    local Shape = require "shape"
    local Circle = setmetatable({}, { __index = Shape })
    Circle.radius = 0.0
    const function Circle:new (x:number, y:number, radius:value):self
      local c:self = setmetatable(Shape:new(x, y), { __index = self })
      c.radius = tonumber(radius) or 0.0
      return c
    end
    const function Circle:area ():number
      return math.pi * self.radius * self.radius
    end
    return Circle
\end{verbatim}

In this example, we use the other two \texttt{setmetatable} rules.
The second rule appears to trigger the refinement of table types,
as we need this rule to add new methods and also to override
existing methods in \texttt{Circle}.
In a \texttt{setmetatable} expression
\texttt{setmetatable(\{\},\{\string_\string_index = id\})},
if \texttt{id} has a \emph{fixed} table type then it is safe to
produce an equivalent \emph{open} table type, as \emph{fixed}
table types do not allow hiding fields.
The third rule appears to redefine the constructor \texttt{new}.
In a \texttt{setmetatable} expression
\texttt{setmetatable(e,\{\string_\string_index = id\})},
if \texttt{e} has a \emph{fixed} table type \texttt{t}, \texttt{id} has
type \texttt{self}, and the type bound to \texttt{self} is a subtype of
\texttt{t}, then the expression has type \texttt{self}.
Note how we can use this rule to call the constructor of \texttt{Shape}
inside the overridden constructor.
A limitation of this class system is that the overridden constructor
must be a subtype of the original constructor,
so the type of \texttt{radius} has to be quite permissive.

Like in the preivous example, the last return statement in the
top-level exports a \emph{fixed} table type, which we could
document through the following class declaration:
\begin{verbatim}
    class Circle
      x, y, radius:number
      const new:(number, number, value) => (self)
      const move:(number, number) => ()
      const area:() => (number)
    end
\end{verbatim}

Note that this class definition introduces a syntactical overhead
to manually include new and overriden methods.
This class definition is equivalent to the alias that exports the
following \emph{fixed} table type to the class \texttt{Circle}:
\begin{verbatim}
    <{ "x":number, "y":number, "radius":number,
       const "new":(self, number, number, value) -> (self),
       const "move":(self, number, number) -> (),
       const "radius":() -> (number) }>
\end{verbatim}

@@

In all examples, if we erase all type and \texttt{const} annotations,
they become valid Lua code, which run with the same semantics as Typed Lua code.
